---
title: "PracticaFinal"
author: "Josep Damià Ruiz Pons"
date: "2023-12-14"
output: html_document
---

```{r}
# Lee las líneas del archivo
lineas <- readLines('FPdata.csv')

# Elimina las comillas al principio y al final de cada línea.
lineas <- gsub("^\"", "", lineas)
lineas <- gsub("\"$", "", lineas)
# Cambia las lineas con doble comilla ("") por una sola (")
lineas <- gsub("\"\"", "\"", lineas)

# Crea un nuevo archivo temporal sin las comillas al principio y al final
# NO SE SI SE PUEDE HACER DE OTRA FORMA SIN EL ARCHIVO TEMPORAL
archivo_temporal <- tempfile(fileext = ".csv")
writeLines(lineas, archivo_temporal)

# Lee el archivo CSV sin comillas al principio y al final
df <- read.csv(file = archivo_temporal, header = TRUE, stringsAsFactors = TRUE, quote = "\"")
```



```{r}
summary(df)
str(df)
```


```{r}
threshold <- 5
rows_with_many_nas <- which(rowSums(is.na(df)) > threshold)
df[rows_with_many_nas, ]
df <- df[-rows_with_many_nas, ]
```

```{r}

# Crear un nuevo data frame con la misma estructura pero sin datos
cdf <- data.frame(df)
```

## State
```{r}
table(df$State)
cdf$State <- factor(df$State)
  # Asigna un número único a cada nivel del factor
  cdf$State <- as.integer(cdf$State)
```

## Gender
```{r}
table(df$Gender, useNA = "ifany")

cdf$Gender <- ifelse(df$Gender == "Female", 0, 1)
table(cdf$Gender)
```
## General health

```{r}
# Convertir el vector de datos a un factor con niveles ordenados de forma ascendente
numeric_general_health <- factor(df$GeneralHealth, levels = c("Poor", "Fair", "Good", "Very good", "Excellent"))
# Asignar -1 a los valores NA utilizando ifelse
cdf$GeneralHealth <- ifelse(is.na(numeric_general_health), 4, as.integer(numeric_general_health))
table(cdf$GeneralHealth)
```

## PhysicalHealthDays

```{r}

PhysicalHealthDays_frq <- table(df$PhysicalHealthDays, useNA = "ifany")
PhysicalHealthDays_frq
mode_PhysicalHealthDays <- names(PhysicalHealthDays_frq[which.max(PhysicalHealthDays_frq)])
mode_PhysicalHealthDays
cdf$PhysicalHealthDays <- ifelse(is.na(df$PhysicalHealthDays), as.numeric(mode_PhysicalHealthDays),as.numeric(df$PhysicalHealthDays))
table(cdf$PhysicalHealthDays)
```
## MentalHealthDays

```{r}

MentalHealthDays_frq <- table(df$MentalHealthDays, useNA = "ifany") 
MentalHealthDays_frq
mode_MentalHealthDays <- names(MentalHealthDays_frq[which.max(MentalHealthDays_frq)])
mode_MentalHealthDays
cdf$MentalHealthDays <- ifelse(is.na(df$MentalHealthDays), as.numeric(mode_MentalHealthDays),as.numeric(df$MentalHealthDays))
table(cdf$MentalHealthDays)
```

## LastCheckupTime
```{r}
table(df$LastCheckupTime, useNA = "ifany") 
# Convertir el vector de datos a un factor con niveles ordenados de forma ascendente
numeric_checkup_time <- factor(df$LastCheckupTime)
# Asignar -1 a los valores NA utilizando ifelse
cdf$LastCheckupTime <- ifelse(is.na(numeric_checkup_time), -1, as.integer(numeric_checkup_time))
table(cdf$LastCheckupTime)
```

## PhysicalActivities

```{r}
table(df$PhysicalActivities, useNA = "ifany") 
cdf$PhysicalActivities <- ifelse(df$PhysicalActivities == "", 1, ifelse(df$PhysicalActivities == "Yes", 1,0))
table(cdf$PhysicalActivities)
```

## SleepHours

```{r}
SleepHours_frq <- table(df$SleepHours, useNA = "ifany") 
SleepHours_frq
mode_SleepHours <- names(SleepHours_frq[which.max(SleepHours_frq)])
mode_SleepHours
cdf$SleepHours <- ifelse(is.na(df$SleepHours), as.numeric(mode_SleepHours),as.numeric(df$SleepHours))
table(cdf$SleepHours)
```
## RemovedTeeth
```{r}
table(df$RemovedTeeth, useNA = "ifany")
cdf$RemovedTeeth <- ifelse(df$RemovedTeeth == "", 1, ifelse(grepl("None", df$RemovedTeeth),0,1))
table(cdf$RemovedTeeth)
```

## HadHeartAttack

```{r}
table(df$HadHeartAttack, useNA = "ifany") 
cdf$HadHeartAttack <- ifelse(df$HadHeartAttack == "", 0, ifelse(df$HadHeartAttack == "No", 0,1))
table(cdf$HadHeartAttack)
```

## HadAngina

```{r}
table(df$HadAngina, useNA = "ifany") 
cdf$HadAngina <- ifelse(df$HadAngina == "", 0, ifelse(df$HadAngina == "No", 0,1))
table(cdf$HadAngina)
```


## HadStroke
```{r}
table(df$HadStroke, useNA = "ifany") 
cdf$HadStroke <- ifelse(df$HadStroke == "", 0, ifelse(df$HadStroke == "No", 0,1))
table(cdf$HadStroke)
```
## HadAsthma
```{r}
table(df$HadAsthma, useNA = "ifany") 
cdf$HadAsthma <- ifelse(df$HadAsthma == "", 0, ifelse(df$HadAsthma == "No", 0,1))
table(cdf$HadAsthma)
```
## HadSkinCancer
```{r}
table(df$HadSkinCancer, useNA = "ifany") 
cdf$HadSkinCancer <- ifelse(df$HadSkinCancer == "", 0, ifelse(df$HadSkinCancer == "No", 0,1))
table(cdf$HadSkinCancer)
```

## HadCOPD
```{r}
table(df$HadCOPD, useNA = "ifany") 
cdf$HadCOPD <- ifelse(df$HadCOPD == "", 0, ifelse(df$HadCOPD == "No", 0,1))
table(cdf$HadCOPD)
```

## HadDepressiveDisorder
```{r}
table(df$HadDepressiveDisorder, useNA = "ifany") 
cdf$HadDepressiveDisorder <- ifelse(df$HadDepressiveDisorder == "", 0, ifelse(df$HadDepressiveDisorder == "No", 0,1))
table(cdf$HadDepressiveDisorder)
```


## HadKidneyDisease
```{r}
table(df$HadKidneyDisease, useNA = "ifany") 
cdf$HadKidneyDisease <- ifelse(df$HadKidneyDisease == "", 0, ifelse(df$HadKidneyDisease == "No", 0,1))
table(cdf$HadKidneyDisease)
```

## HadArthritis
```{r}
table(df$HadArthritis, useNA = "ifany") 
cdf$HadArthritis <- ifelse(df$HadArthritis == "", 0, ifelse(df$HadArthritis == "No", 0,1))
table(cdf$HadArthritis)
```

## HadDiabetes
```{r}
table(df$HadDiabetes, useNA = "ifany") 
cdf$HadDiabetes <- ifelse(df$HadDiabetes == "", 0, ifelse(grepl("No",df$HadDiabetes), 0,1))
table(cdf$HadDiabetes)
```

## DeafOrHardOfHearing
```{r}
table(df$DeafOrHardOfHearing, useNA = "ifany") 
cdf$DeafOrHardOfHearing <- ifelse(df$DeafOrHardOfHearing == "", 0, ifelse(df$DeafOrHardOfHearing == "No", 0,1))
table(cdf$DeafOrHardOfHearing)
```

## BlindOrVisionDifficulty
```{r}
table(df$BlindOrVisionDifficulty, useNA = "ifany") 
cdf$BlindOrVisionDifficulty <- ifelse(df$BlindOrVisionDifficulty == "", 0, ifelse(df$BlindOrVisionDifficulty == "No", 0,1))
table(cdf$BlindOrVisionDifficulty)
```

## DifficultyConcentrating
```{r}
table(df$DifficultyConcentrating, useNA = "ifany") 
cdf$DifficultyConcentrating <- ifelse(df$DifficultyConcentrating == "", 0, ifelse(df$DifficultyConcentrating == "No", 0,1))
table(cdf$DifficultyConcentrating)
```
## DifficultyWalking
```{r}
table(df$DifficultyWalking, useNA = "ifany") 
cdf$DifficultyWalking <- ifelse(df$DifficultyWalking == "", 0, ifelse(df$DifficultyWalking == "No", 0,1))
table(cdf$DifficultyWalking)
```
## DifficultyDressingBathing
```{r}
table(df$DifficultyDressingBathing, useNA = "ifany") 
cdf$DifficultyDressingBathing <- ifelse(df$DifficultyDressingBathing == "", 0, ifelse(df$DifficultyDressingBathing == "No", 0,1))
table(cdf$DifficultyDressingBathing)
```

## DifficultyErrands
```{r}
table(df$DifficultyErrands, useNA = "ifany") 
cdf$DifficultyErrands <- ifelse(df$DifficultyErrands == "", 0, ifelse(df$DifficultyErrands == "No", 0,1))
table(cdf$DifficultyErrands)
```

## SmokerStatus
```{r}
table(df$SmokerStatus, useNA = "ifany") 
cdf$SmokerStatus <- ifelse(df$SmokerStatus == "", 0, ifelse(grepl("Never smoked",df$SmokerStatus), 0,1))
table(cdf$SmokerStatus)
```

## ECigarretteUsage
```{r}
table(df$ECigaretteUsage, useNA = "ifany") 
cdf$ECigaretteUsage <- ifelse(df$ECigaretteUsage == "", 0, ifelse(grepl("Never smoked",df$ECigaretteUsage) | grepl("Not at all",df$ECigaretteUsage) , 0,1))
table(cdf$ECigaretteUsage)
```

## ChestScan
```{r}
table(df$ChestScan, useNA = "ifany") 
cdf$ChestScan <- ifelse(df$ChestScan == "", 0, ifelse(df$ChestScan == "No", 0,1))
table(cdf$ChestScan)
```


## RaceEthnicityCategory
```{r}
cdf$RaceEthnicityCategory <- df$RaceEthnicityCategory
table(df$RaceEthnicityCategory, useNA = "ifany") 

cdf$RaceEthnicityCategory[df$RaceEthnicityCategory == ""] <- "White only, Non-Hispanic"
cdf$RaceEthnicityCategory <- factor(cdf$RaceEthnicityCategory)
  cdf$RaceEthnicityCategory <- as.integer(cdf$RaceEthnicityCategory)
  table(cdf$RaceEthnicityCategory)
```

## AgeCategory
```{r}
table(df$AgeCategory, useNA = "ifany") 
cdf$AgeCategory[df$AgeCategory == ""] <- "Age 65 to 69"
cdf$AgeCategory <- factor(cdf$AgeCategory)
  cdf$AgeCategory <- as.integer(cdf$AgeCategory)
  table(cdf$AgeCategory)
```

## HeightInMeters

```{r}
height_frq <- table(df$HeightInMeters, useNA = "ifany") 
height_frq
mode_height <- names(height_frq[which.max(height_frq)])
cdf$HeightInMeters <- ifelse(is.na(df$HeightInMeters),as.numeric(mode_height),as.numeric(df$HeightInMeters) )
table(cdf$HeightInMeters, useNA = "ifany") 
```
## WeightInKilograms



```{r}
weight_frq <- table(df$WeightInKilograms) 
weight_frq
mode_weight <- names(weight_frq[which.max(weight_frq)])
mode_weight
cdf$WeightInKilograms <- ifelse(is.na(df$WeightInKilograms),as.numeric(mode_weight),as.numeric(df$WeightInKilograms) )
table(cdf$WeightInKilograms, useNA = "ifany") 

```
## BMI
```{r}
BMI_frq <- table(df$BMI) 
mode_BMI <- names(BMI_frq[which.max(BMI_frq)])
mode_BMI
cdf$BMI <- ifelse(is.na(df$BMI),as.numeric(mode_BMI),as.numeric(df$BMI) )
hist(cdf$BMI)
```



```{r}
quantiles = quantile(cdf$BMI)
q1 <- quantiles[2]
q3 <- quantiles[4]

iqr = q3 - q1
lower_bound = q1 - 1.5 * iqr
upper_bound = q3 + 1.5 * iqr

outliers <- cdf$BMI[cdf$BMI < lower_bound | cdf$BMI > upper_bound]
cdf[cdf$BMI < lower_bound | cdf$BMI > upper_bound, ]
df <- cdf[!(cdf$BMI < lower_bound | cdf$BMI > upper_bound), ]
cdf <- cdf[!(cdf$BMI < lower_bound | cdf$BMI > upper_bound), ]
```

```{r}
hist(cdf$BMI)
table(cdf$HeightInMeters)
hist(cdf$HeightInMeters)
```

## AlcoholDrinkers

```{r}

table(cdf$AlcoholDrinkers, useNA = "ifany") 
cdf$AlcoholDrinkers <- ifelse(cdf$AlcoholDrinkers == "", 1, ifelse(cdf$AlcoholDrinkers == "No", 0,1))
table(cdf$AlcoholDrinkers)
```

## HIVTesting

```{r}
table(cdf$HIVTesting, useNA = "ifany") 
cdf$HIVTesting <- ifelse(cdf$HIVTesting == "", 0, ifelse(cdf$HIVTesting == "No", 0,1))
table(cdf$HIVTesting)
```

## FluVaxLast12

```{r}
table(cdf$FluVaxLast12, useNA = "ifany") 
cdf$FluVaxLast12 <- ifelse(cdf$FluVaxLast12 == "", 1, ifelse(cdf$FluVaxLast12 == "No", 0,1))
table(cdf$FluVaxLast12)
```

## PneumoVaxEver

```{r}
table(cdf$PneumoVaxEver, useNA = "ifany") 
cdf$PneumoVaxEver <- ifelse(cdf$PneumoVaxEver == "", 0, ifelse(cdf$PneumoVaxEver == "No", 0,1))
table(cdf$PneumoVaxEver)
```

## TetanusLast10Tdap

```{r}
tetanus_frq <- table(cdf$TetanusLast10Tdap, useNA = "ifany") 
tetanus_frq
mode_TetanusLast10Tdap <- names(tetanus_frq[which.max(tetanus_frq)])
mode_TetanusLast10Tdap
cdf$TetanusLast10Tdap <- ifelse(cdf$TetanusLast10Tdap == "",0,ifelse(grepl(cdf$TetanusLast10Tdap, "No"), 0, 1) )
table(cdf$TetanusLast10Tdap, useNA = "ifany")
```

## HighRiskLastYear

```{r}
table(cdf$HighRiskLastYear, useNA = "ifany") 
cdf$HighRiskLastYear <- ifelse(cdf$HighRiskLastYear == "", 0, ifelse(cdf$HighRiskLastYear == "No", 0,1))
table(cdf$HighRiskLastYear)
```

## CovidPos

```{r}
cdf$CovidPos <- df$CovidPos
table(df$CovidPos, useNA = "ifany") 
cdf$CovidPos <- ifelse(df$CovidPos == "",0,ifelse(grepl("No", df$CovidPos), 0, 1))
table(cdf$CovidPos, useNA = "ifany")
```

```{r}
str(cdf)
```
```{r}
heatmap(cor(cdf))
```

## Question two: Can we predict the general health with the available data?

```{r}
normalize <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}

columns_to_normalize <- c("PhysicalHealthDays", "MentalHealthDays", "SleepHours")
cdf[columns_to_normalize] <- lapply(cdf[columns_to_normalize], normalize)
cdf[columns_to_normalize] <- lapply(cdf[columns_to_normalize], function(x) round(x, 2))
```

Classes are balanced ?
```{r}
# Assuming 'cdf' is your dataset and 'GeneralHealth' is the column containing numeric class labels (1, 2, 3, 4, 5)
barplot(table(cdf$GeneralHealth), col = "skyblue", main = "Non-balanced Class Distribution")

# Calculate label frequencies
label_frequencies <- table(cdf$GeneralHealth)

# Find the label with the minimum frequency
min_freq_label <- as.numeric(names(label_frequencies[label_frequencies == min(label_frequencies)]))

# Get the minimum frequency count
min_freq_count <- min(label_frequencies)

# Balance classes by selecting the same number of samples for each label
balanced_data <- do.call(rbind, lapply(unique(cdf$GeneralHealth), function(label) {
  subset(cdf, GeneralHealth == label)[sample(nrow(subset(cdf, GeneralHealth == label)), min_freq_count), ]
}))

# Display the balanced class distribution
barplot(table(balanced_data$GeneralHealth), col = "skyblue", main = "Balanced Class Distribution")

```


```{r}
set.seed(43)

ind <- sample(2, nrow(cdf), replace=TRUE, prob=c(0.80, 0.20))

train_indices <- sample(1:nrow(cdf), 0.75 * nrow(cdf))
train_data <- cdf[ind==1,]
test_data <- cdf[ind==2,]

train_data <- balanced_data
#test_data <- test_data[1:10000, ]
```

KNN 

```{r}
library(gmodels)
library(class)

predictor_cols <- setdiff(colnames(df), "GeneralHealth")

calssifiers.KNN <- knn(train = train_data[, predictor_cols], 
           test = test_data[, predictor_cols], 
           cl = train_data$GeneralHealth, 
           k = 6)

results <- CrossTable(x = test_data$GeneralHealth, y = calssifiers.KNN, prop.chisq = FALSE)
accuracy.KNN <- sum(results$t[1:2, 1:2]) / sum(results$t)
print(paste("Knn accuracy:", round(accuracy.KNN * 100, 2), "%"))
```

Naïve

```{r}
library(e1071)

classifiers.NB <- naiveBayes(x = train_data[, predictor_cols], y = train_data$GeneralHealth)
predicts.NB <- predict(classifiers.NB, newdata = test_data[, predictor_cols])
results.NB <- data.frame(Actual = test_data$GeneralHealth, Predicted = predicts.NB)
accuracy.NB <- sum(results.NB$Actual == results.NB$Predicted) / nrow(results.NB)
print(paste("Naive Bayes accuracy:", round(accuracy.NB * 100, 2), "%"))
```

---

Decision trees:

```{r}
library(RWeka)
library(party)
library(partykit)
library(FSelector)
```

```{r}
classifiers.ID3 <- J48(as.factor(GeneralHealth) ~ ., data = train_data)
summary(classifiers.ID3)
```

```{r}
predict_model<-predict(classifiers.ID3, test_data) 
m_at <- table(as.factor(test_data$GeneralHealth), predict_model) 
print(m_at)
ac_Test <- sum(diag(m_at)) / sum(m_at)
print(paste('Accuracy for test is found to be', ac_Test))
```


```{r}
library(C50)

classifiers.C50 <- C5.0(as.factor(GeneralHealth) ~ ., data = train_data)

#Asset acurracy
predict_model<-predict(classifiers.C50, test_data) 
m_at <- table(as.factor(test_data$GeneralHealth), predict_model) 
print(m_at)
ac_Test <- sum(diag(m_at)) / sum(m_at)
print(paste('Accuracy for test is found to be', ac_Test))
```

```{r}
library(rpart.plot)
# Assuming 'train_data' is your dataset and 'GeneralHealth' is the column containing class labels

# Build the rpart model
classifier_rpart <- rpart(GeneralHealth ~ ., data = train_data, method = "class", control = rpart.control(cp = 1e-5))

bestcp <- classifier_rpart$cptable[which.min(classifier_rpart$cptable[,"xerror"]),"CP"]

# Step3: Prune the tree using the best cp.
classifier_rpart.pruned <- prune(classifier_rpart, cp = bestcp)

# Make predictions on the training data
train_pred_rpart <- predict(classifier_rpart.pruned, train_data[, predictor_cols], type = 'class')

# Compare predicted and actual values
accuracy_rpart <- sum(train_pred_rpart == train_data$GeneralHealth) / nrow(train_data)

# Print the accuracy
print(paste('Accuracy for training data using rpart is found to be', round(accuracy_rpart * 100, 2), '%'))
```


---

```{r}
library(ranger)

# Build the Random Forest model
ranger <- ranger(
    formula   = as.factor(GeneralHealth) ~ ., 
    data      = train_data, 
    num.trees = 5000,
    mtry = sqrt(ncol(train_data)), # You can adjust the number of features considered for each split
    importance = "permutation"
  )

```

```{r}
# Make predictions on the test data
test_pred_rf <- predict(ranger, data = test_data)$predictions

# Compare predicted and actual values
accuracy_rf <- sum(test_pred_rf == test_data$GeneralHealth) / nrow(test_data)

# Print the accuracy
print(paste('Accuracy for test data using ranger is found to be', round(accuracy_rf * 100, 2), '%'))

```

