---
title: "Task1 - Mean, Variation and Covariance"
author: "Martí Paredes Salom"
date: '2023-09-16'
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

In this first task of the assignment data mining we need to calculate the mean and variation of a data set, in this case we will be using the RStudio built-in data set `ChickWeight`.

First of all we will have a look at the data set, using the heads command to show the attributes available for each chick.

```{r}
df <- ChickWeight
head(df)
```

In this data set we see that there are **4 columns**: Weight, time, chick and diet, in this first exercise we will center in the **first two columns**.

### Mean and variation

As said, the mean and variation. with the following formula:

$$\overline{x} = \frac{\sum_{i=1}^n x_i}{n}$$ $$s_x^2 = \frac{\sum_{i=1}^n (x_i-\overline{x})^2}{n-1}$$

```{r}
meanChicksWeigth <- sum(df$weight) / length(df$weight) 
variationChicksWeigth <- sum((df$weight - meanChicksWeigth)^2)/(length(df$weight) -1)
```

Done the calculations the **mean of the weight** we obtained a value of **121.82** and a **variation of 5051.22**, now we will do the same with the height.

```{r}
meanChicksTime <- sum(df$Time) / length(df$Time) 
variationChicksTime <- sum((df$Time - meanChicksTime)^2)/(length(df$Time) -1)
```

This time we obtain that the **mean of time is 10.72** and the **variation is 45.68**.

### Covariance

Finally we will calculate the covariance between the weight and time, to do it we will use the following formula:

$$ s_{xy} = \frac{\sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}{n-1} $$

```{r}
covWeigthTime <- sum((df$weight - meanChicksWeigth)*(df$Time - meanChicksTime)) / (length(df$weight) - 1)
```

<!-- We obtained a **covariance of 401.39**, a positive covariance indicates that as the weight of the chicks increases, the 'Time' it takes for them to reach a certain growth stage also tends to increase. In other words, there is a positive linear relationship between 'weight' and 'Time,' suggesting that heavier chicks generally take more time to reach the specified growth stage.  -->
